---
title: "PA1_template.Rmd"
author: "E.D."
date: "October 20, 2016"
output: html_document
---

```{r}

###################################################################################################
## E.D. October 2016
## DS 5 Reproducible Research - Project 1

###################################################################################################
## 1.	Code for reading in the dataset and/or processing the data

## Download the data file.
if(!file.exists("./mydata_DS5")){dir.create("./mydata_DS5")}
 fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
 download.file(fileUrl,destfile="./mydata_DS5/Dataset.zip")

## Unzip the dataSet to /mydata_DS5 directory
 unzip(zipfile="./mydata_DS5/Dataset.zip",exdir="./mydata_DS5")

## Read in our data set
ActivityData <- read.csv('./mydata_DS5/activity.csv',stringsAsFactors = FALSE)

###################################################################################################
## 2.	Histogram of the total number of steps taken each day

## Generate a set of unique days
DaysSet = unique(ActivityData$date)

## Initialize Counters 
TotalStepsEachDay = 0  
MeanStepsEachDay = 0  
MaxStepsEachDay = 0  
MaxStepsInterval = vector(mode="numeric", length=length(DaysSet)) 

## Examine each day, traverse the set.
for (i in 1:length(DaysSet)) {
  ## Temporary day
  CurrentDay <- DaysSet[i]
  ## Find index of intervals matching temporary day
  CurrentDayIn <-ActivityData$date == CurrentDay
  ## Use index to get step values and traverse.
  GetStepValues <- ActivityData$steps[CurrentDayIn]
  ## Find the maximum value for all intervals during a temporary day
  MaxStepsEachDay[i] <- max(GetStepValues)
  
  ## Fill in the missing Maximum with NA
  if (is.na(MaxStepsEachDay[i])){
    MaxStepsInterval[i] = NA
  }  else {
    ## Else find the interval of maximum steps
    MaxStepsInterval[i] <- which.max(GetStepValues)
  }
  
  ## Find the total number of steps per day  
  TotalStepsEachDay[i] = sum(GetStepValues)
  ## Get average number of steps per interval per day
  MeanStepsEachDay[i] = mean(GetStepValues, na.rm = TRUE)
}

## Setup a png file to prepare for a plot.
#png("Plot2RR.png", width=480, height=480) 

# Plot histogram of total number of steps per day
hist(TotalStepsEachDay, breaks = 5, col="blue", main='Total Number of Steps Each Day',
     ylab='5 Minutes Interval', xlab='# Steps per Day')

# Close the file.
#dev.off

###################################################################################################
## 3.	Mean and median number of steps taken each day
meanPerDayt = mean(TotalStepsEachDay, na.rm = TRUE)
print(paste("The Mean Per Day is",as.character(round(meanPerDayt)),sep = " "))
```
[1] "The Mean Per Day is 10766"
``` {r}
medPerDayt = median(TotalStepsEachDay, na.rm = TRUE)
print(paste("The Median Per Day is",as.character(round(medPerDayt)),sep = " "))
```
[1] "The Median Per Day is 10765""
```{r}
###################################################################################################
## 4.	Time series plot of the average number of steps taken

## Compute the index for days with NA
DaysSet = unique(ActivityData$date)

DayList = vector(mode="numeric", length=0)

## Traverse the Days Set
for (i in 1:length(DaysSet)) {
  
  CurrentDay <- DaysSet[i]
  CurrentDayIn <- ActivityData$date == CurrentDay
  GetStepValues <- ActivityData$steps[CurrentDayIn]
  DayList = cbind(DayList, GetStepValues) }

  meanInt = round(rowMeans(DayList,na.rm = TRUE))

## Setup a png file to prepare for a plot.
#png("Plot4RR.png", width=480, height=480) 

plot(meanInt,type = 'l',xlab = '5 Minute Interval',ylab = 'Average Across All Days', main='Average Daily Activity Pattern')

# Close the file.
#dev.off()
```
```{r}
###################################################################################################
## 5.	The 5-minute interval that, on average, contains the maximum number of steps
mean5minInt <- round(mean(MaxStepsInterval, na.rm = TRUE))
print(paste(as.character(mean5minInt)," is the 5-minute interval that, on average, contains the max number of steps",sep = " "))
## [1] "145  is the 5-minute interval that, on average, contains the max number of steps"

###################################################################################################
## 6.	Code to describe and show a strategy for imputing missing data

## Algorithm:
## Initialize a maxtrix (r,c), where r=intervals and c= days.
## Look at each row of the matrix that has an interval for all days.
## Determine the index of the days with missing days for a particular interval.
## Calculate the average of days with intervals present
## Replace the intervals of NA, with the average of the intervals that possess data.

# Unique days
DaysSet = unique(ActivityData$date)
# Initialize the vector
DayList = vector(mode="numeric", length=0)

# Traverse the Days Set
for (i in 1:length(DaysSet)) {
  # Get temporary day
  CurrentDay <- DaysSet[i]
  # Get index for temporary day
  CurrentDayIn <- ActivityData$date == CurrentDay
  # Use index to get interval step data
  GetStepValues <- ActivityData$steps[CurrentDayIn]
  # Insert interval data into column of matrix
  DayList = cbind(DayList, GetStepValues)  }

## Number of missing intervals
numMissintervals = sum(is.na(DayList))
print(paste("The total number of missing intervals is ",as.character(numMissintervals),sep = " "))
```
[1] "The total number of missing intervals is  2304"
```{r}
# Initialize an empty matrix
DayList2 = matrix(data=NA,nrow=dim(DayList)[1],ncol=dim(DayList)[2])

# Traverse the Days List
for (i in 1:dim(DayList)[1]) {
  
  tempRow = DayList[i,]
  # Find the index of missing values
  naInd = is.na(tempRow)
  # Get mean of values present
  mRow = mean(tempRow, na.rm = TRUE)
  # Replace missing values with average
  tempRow[naInd] = round(mRow)
  # Insert new row into the matrix
  DayList2[i,] = tempRow    }

meanPerDay2 = mean(colSums(DayList2))
print(paste("The Mean Per Day with no missing values is",as.character(round(meanPerDay2)),sep = " "))
```
[1] "The Mean Per Day with no missing values is 10766"
```{r}
medPerDay2 = median(colSums(DayList2))
print(paste("The Median Per Day with no missing values is",as.character(round(medPerDay2)),sep = " "))
```
[1] "The Median Per Day with no missing values is 10762"

The data seems to be fairly similar as with the imputed missing values.

```{r}
###################################################################################################
## 7.	Histogram of the total number of steps taken each day after missing values are imputed
# Get unique days
DaysSet = unique(ActivityData$date)
# Preallocate empty data holder
totPerDay2 = 0
# Loop through days
for (i in 1:length(DaysSet)) {
  # Get temporary column = intervals for single day with missing values replaced by means
  tempCol = DayList2[,i]
  # Get new sum with imputed missing values
  totPerDay2[i] = sum(tempCol)
  
}

## Setup a png file to prepare for a plot.
#png("Plot7RR.png", width=480, height=480) 

# Generate the histogram
hist(totPerDay2, breaks = 10, main = 'Total after missing values imputed',
     ylab = '5 Minutes Interval', xlab = 'Total Number of Steps', col='blue')

# Close the file.
#dev.off()

###################################################################################################
# 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
# Get list of unique days
DaysSet = unique(ActivityData$date)

# Initialize vector for weekend day information
weekEndDays = vector(mode="numeric", length=0)
# initialize vector for week day information
weekDays = vector(mode="numeric", length=0)

# Traverse the set
for (i in 1:length(DaysSet)) {
  
  CurrentDay <- DaysSet[i] 
  
  CurrentDayIn <- ActivityData$date == CurrentDay
  
  GetStepValues <- ActivityData$steps[CurrentDayIn] 
  
  ## All intervals are missing skip this day
  if (sum(!is.na(GetStepValues))==0) {
    next
  } else {
    # Check whether the day is a weekday or weekend day
    checkDay = weekdays(as.Date(CurrentDay))
    
    # Determine if the day is in the weekday set or weekend set.
    if (checkDay == 'Saturday' | checkDay == 'Sunday'){
      weekEndDays = cbind(weekEndDays, GetStepValues)
      ## The day is a weekday (Monday, Tuesday, Wednesday, Thursday, Friday)
    } else {
      weekDays = cbind(weekDays, GetStepValues)
    }
  }
}

## Calculate the mean of the rows of the weekend matrix
weekEndMeans <- round(rowMeans(weekEndDays))
## Calculate the mean of the rows of the week matrix
weekMeans <- round(rowMeans(weekDays))

## Generate the plots for the weekend and weekdays.
par(mfrow = c(1,2))

## Setup a png file to prepare for a plot.
#png("Plot8aRR.png", width=480, height=480) 

plot(weekEndMeans,type = "l",  xlab = '5 Minutes Interval', ylab = 'Average Number of Steps', 
     main = 'Activity Pattern Weekend')

## Close the file for the weekend plot.
##dev.off()

#png("Plot8bRR.png", width=480, height=480) 
plot(weekMeans,type = "l", xlab = '5 Minutes Interval', ylab = 'Average Number of Steps',
     main = 'Activity Pattern Weekdays')
## Close the file for the weekday plot.
#dev.off()

###################################################################################################
```
There does seem to be some difference between the weekdays and weekend trends. Most people have more free
time on the weekend, and thus probably fit in some more exercise time. On the other hand,during the week when most people 
work they may be tired, or have other obligations, and will execercise less during the week.
